A Logical Solution of the Linearity Challenge

### 2025-07-31 paper/calculus.tex
agda code
better definitions

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) peer endpoints are always used in complementary ways. This
latter property is dubbed \emph{well formedness} in the text of the challenge.
We now formulate these properties using the notation developed so far.

Concerning the preservation of typing, this corresponds to the usual subject
reduction result which is expressed thus.

@@

In order to formulate well formedness, we start by defining \emph{reduction
contexts} are partial processes with a single unguarded hole $\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
\]

As usual, we write $\ReductionContext[P]$ for the process obtained by replacing
the hole in $\ReductionContext$ with $P$, noting that such replacement may
capture channel names that are bound in $\ReductionContext$ and occur free in
$P$.

Next we introduce some terminology concerning the shape of processes. We say
that $P$ is a \emph{thread} if it is anything but a cut. In other words, a
thread is either a link or a process that starts with an input/output action of
some sort. Note that a thread may contain cuts, but these must all guarded
underneath the topmost action prefix.
%
We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
An observable process is a process that exposes an action on a free channel name
and therefore can interact through that channel, if put into some appropriate
reduction context. We say that $P$ is \emph{reducible} if $P \red Q$ for some
$Q$. A reducible process may perform a reduction step.

Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
complementary ways, then their parallel composition $\Cut\x{P}{Q}$ must be able
to reduce according to one of the principal cut reductions described in
\Cref{tab:semantics}. Therefore, an alternative (and more convenient) way of
formulating well formedness is the following:

\begin{definition}[well formedness]
    \label{def:well-formedness}
    We say that $P$ is \emph{well formed} if $P \pcong
    \ReductionContext[\Cut\x{Q_1}{Q_2}]$ where both $Q_1$ and $Q_2$ are threads
    implies that $\Cut\x{Q_1}{Q_2}$ is either observable or reducible.
\end{definition}

In general a process may be well-formed even if it is unable to reduce. For
example, $P \eqdef \Cut\x{\Wait\y.\Close\x}{\Wait\x.\Close\z}$ is irreducible
but also well formed, because $P \pcong
\Wait\y.\Cut\x{\Close\x}{\Wait\x.\Close\z}$ which is observable.

@@

% \begin{theorem}[deadlock freedom]
%     \label{thm:df}
%     If $\wtp{P}\Context$, then $P$ is either observable or reducible.
% \end{theorem}

% safety

>>>>

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) peer endpoints are always used in complementary ways. This
latter property is dubbed \emph{well formedness} in the text of the challenge.
In this work we also consider \emph{deadlock freedom} which follows from the
fact that \Calculus is based on linear logic. As we will see, well formedness is
nothing but a particular form of deadlock freedom. We now formulate these
properties using the notation developed so far.

Concerning the preservation of typing, this corresponds to the usual subject
reduction result, which is expressed thus:

@@

In order to formulate deadlock freedom, we need to introduce some terminology
for referring to processes that are unable to make any progress. Deadlock is not
simply the negation of reducibility: there are irreducible processes that are
not a deadlock because they would be able to reduce if put in a suitable
context. For example, $\Wait\x.P$ does not reduce, and yet it is not a deadlock
because it would be able to make progress when composed in parallel with
$\Close\x$. In general, even a process like $\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where
$x\ne y,z$ cannot be considered a deadlock, because the prefixes $\Wait\y$ and
$\Wait\z$ could be exposed using \SWait and possibly \SComm. Let us make all
this more precise.

We say that $P$ is a \emph{thread} if it is anything but a cut. In other words,
a thread is either a link or a process that starts with an input/output action
of some sort. Note that a thread may contain cuts, but these must all guarded
underneath the topmost action prefix.
%
We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
An observable process is a process that exposes an action on a free channel name
and therefore can interact through that channel, if put into some appropriate
context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
reducible process may perform a reduction step.
%
Now, a process is a \emph{deadlock} if it is neither observable nor reducible. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$.

\begin{theorem}[deadlock freedom]
    \label{thm:df}
    If $\wtp{P}\Context$, then $P$ is not a deadlock.
\end{theorem}

In order to formulate \emph{well formedness}~\cite{CarboneEtAl}, we define
\emph{reduction contexts} as partial processes with a single unguarded hole
$\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
\]

As usual, we write $\ReductionContext[P]$ for the process obtained by replacing
the hole in $\ReductionContext$ with $P$, noting that such replacement may
capture channel names that are bound in $\ReductionContext$ and occur free in
$P$.
%
Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
complementary ways, then their parallel composition $\Cut\x{P}{Q}$ is reducible
according to one of the principal cut reductions described in
\Cref{tab:semantics}. Therefore, an alternative (and more general) way of
formulating well formedness is this:
%
we say that $P$ is \emph{well formed} if $P \pcong
\ReductionContext[\Cut\x{Q_1}{Q_2}]$ implies that $\Cut\x{Q_1}{Q_2}$ is not a
deadlock.

@@

Note that the properties expressed in \Cref{thm:df,thm:type-safety} are
invariant under reductions thanks to \Cref{thm:red}.
